Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-03 tex/manuscript.tex
Submitted version
mer fikling

% Manuscript version (date) in separate footnote
\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version \today.}

@@

A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}). This allows the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora. ComBat has thus helped popularise this approach. Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported. The R package \texttt{limma} (\citealp{Smyth2003}) also offers batch adjustment with covariates, but notes that this is not intended used prior to linear modeling.

@@

In an experiment by \citet{Towfic2014}, cells were treated with either glatiramer acetate (a medicine for multiple sclerosis) or a generic drug, and mRNA was measured using microarrays alongside control sample measurements. A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data were tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}). Unfortunately, the batch--treatment design was unbalanced, with several batches having only one of the main treatments of interest. When we re-analyzed their data without the use of ComBat, but instead blocked for batch effect in \texttt{limma}, only 9 genes were found ($\text{FDR}<0.05$). The sanity check  outlined above with random numbers was also carried out. The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.  Our conclusion is that most of the genes reported as differentially expressed in \citet{Towfic2014} are false positives. This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference.

@@

The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2''). The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up containing 6+2, 3+4 and 9+6 samples. ComBat was applied followed by a T-test, in order to identify differentially expressed genes. First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the T-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$). In addition, the sanity checks outlined above were performed. The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b. In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis do not indicate a huge difference. Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.

@@

A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent. Such tools include GenePattern (\citealp{Reich2006}), which offers ComBat with covariates, as well as TCGA and  inSilicoDb (\citealp{Taminau2011}), presently without covariates as an option.

@@

The data and scripts used to generate the results in this work are available at
\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}.
Additional analyses, performed, but not included in the article, may be found at
\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}.

>>>>

% Manuscript version (date) in separate footnote
%\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version \today.}

@@

A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}). This allows the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora. ComBat has thus helped popularise this approach. Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported. The commercial software Partek (Partek® Genomics Suite®, version 6.6 Copyright ©; 2014 Partek Inc., St. Louis, MO, USA), and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{http://link.springer.com/article/10.1007/s12561-013-9081-1}) also offer batch adjustment with covariates. Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to focus more on the popular and well documented ComBat.

@@

In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug. Cells were treated with
glatiramer acetate (34 samples), the generic drug (11 samples) or 14 different types of other treatments(64 samples), and mRNA was measured using microarrays. A batch effect correlating to the chip (Illumina
WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with
ComBat, whereafter the data were tested for differentially expressed
genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}).
Unfortunately, the batch--treatment design was unbalanced, with several
batches having only one of the main treatments of interest. When we
re-analyzed the data without the use of ComBat, but instead blocked for
batch effect in \texttt{limma}, only 9 genes were found
($\text{FDR}<0.05$). The sanity check  outlined above with random numbers
was also carried out. The distribution of p-values for different settings
are shown in Figure~\ref{fig:comparison}a. Our conclusion is that most of
the genes found as differentially expressed in the analysis described in
\citet{Towfic2014} are false positives. This example represents somewhat
of a ``worst case'' scenario for applying ComBat, since it both has a
very unbalanced batch--group design and an a priori assumption of no
difference. Our re-analysis, although based on the "methods" section of
Towfic et al, deviates in important ways from the analysis that was
actually performed in the paper (personal communication, Ben Zeskind).
More detailed information on this aspect can be found in the extended github repostiory

@@

The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2''). The experiment consist of 30 samples in 3 batches (batch1:6/2, batch2:3/4 and batch3:9/6 treatment/control samples). ComBat was applied followed by a T-test, in order to identify differentially expressed genes. First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the T-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$). In addition, the sanity checks outlined above were performed. The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b. In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis do not indicate a huge difference. Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.

@@

A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent. Such tools include GenePattern (v3.9.0) (\citealp{Reich2006}), AltAnalyze (v2.0.8 http://nar.oxfordjournals.org/content/38/suppl_2/W755.full) and SCAN.UPC (v2.6.3 http://www.pnas.org/content/110/44/17778.long), which offers ComBat with covariates, as well as TCGA and  inSilicoMerging (v1.8.7) (\citealp{ http://www.biomedcentral.com/1471-2105/13/335}), presently without covariates as an option. Especially troublesome, are the ComBat implementations in ChAMP (v1.2.8) (\citealp{http://bioinformatics.oxfordjournals.org/content/30/3/428.full}) and intCor (v1.03) (http://bioinformatics.oxfordjournals.org/content/early/2014/07/24/bioinformatics.btu423.full), where the use of covariates seems automatic and hidden from the user in the method call, while there is no mention of covariates in the documentation or articles.

@@

The data and scripts used to generate the results in this work are available at
\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}.
Additional analyses, performed, but not included in the article, may be found at the extended repository 
\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}.
